---
title: "Importance Sampling for Thomas Process: Simulation Study Report"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
    df_print: kable
params:
  study_file: "study2.RDa"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

required_pkgs <- c("plotly", "DT")
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop("Missing package(s): ", paste(missing_pkgs, collapse = ", "),
       ". Install with: install.packages(c(", paste0('"', missing_pkgs, '"', collapse = ", "), "))")
}
library(plotly)
library(DT)
```

```{r load-data}
# To look at a different study file, either edit the `study_file` param above,
# or render with e.g.:
#   rmarkdown::render("is_thomas_report.Rmd", params = list(study_file = "study1.RDa"))
load(params$study_file)
```

This report summarizes `r params$study_file`: `r length(simStudyResults)` parameter-goal group(s), each with `r length(simStudyResults[[1]])` reweighting scenario(s) (a baseline simulated at `par_0`, reweighted via importance sampling to estimate at `par_goal`).

```{r helpers}
# Same K-function contrast metric as in is_thomas_analysis.R: sum of squared
# differences between K_true^q and K_estimate^q, integrated over r.
calc_K_diff <- function(K_true, K_norm, K_unnorm, q = 1/4) {
  dr <- K_true$r[2] - K_true$r[1]
  diff_norm <- numeric(length(K_norm))
  diff_unnorm <- numeric(length(K_unnorm))
  for (i in seq_along(K_norm)) {
    diff_norm[i] <- sum((K_true$border^q - K_norm[[i]]$border^q)^2) * dr
    diff_unnorm[i] <- sum((K_true$border^q - K_unnorm[[i]]$border^q)^2) * dr
  }
  list(normalized = diff_norm, unnormalized = diff_unnorm)
}

# Human-readable label for a case's delta = par_0 - par_goal, e.g. "kappa+0.20, mu-0.20"
describe_case <- function(case) {
  delta <- case$par_0 - case$par_goal
  parts <- character(0)
  for (nm in names(delta)) {
    v <- delta[[nm]]
    if (abs(v) > 1e-8) parts <- c(parts, sprintf("%s%+.2f", nm, v))
  }
  if (length(parts) == 0) return("par_0 = par_goal")
  paste(parts, collapse = ", ")
}

format_params <- function(p) sprintf("kappa=%.2f, omega=%.2f, mu=%.2f", p$kappa, p$omega, p$mu)
```

## Summary table

One row per (group, scenario), using the largest `nSamples` available in each scenario. Sorted by group; click column headers to sort/search. Rows with >10% relative error in the normalized intensity estimate are highlighted.

```{r summary-table}
rows <- list()
for (g in seq_along(simStudyResults)) {
  group <- simStudyResults[[g]]
  for (i in seq_along(group)) {
    case <- group[[i]]
    n_last <- length(case$res_is$rho_is)

    K_diff <- calc_K_diff(K_true = case$res_true$K_true,
                           K_norm = case$res_is$K_is_norm,
                           K_unnorm = case$res_is$K_is)

    rho_true <- case$res_true$rho_true
    rho_u <- case$res_is$rho_is[n_last]
    rho_n <- case$res_is$rho_is_norm[n_last]

    rows[[length(rows) + 1]] <- data.frame(
      Group = g,
      Scenario = i,
      Delta = describe_case(case),
      Goal = format_params(case$par_goal),
      n_max = case$nSamples[n_last],
      rho_true = rho_true,
      rho_unnorm = rho_u,
      rho_norm = rho_n,
      rho_relerr_unnorm_pct = 100 * (rho_u - rho_true) / rho_true,
      rho_relerr_norm_pct = 100 * (rho_n - rho_true) / rho_true,
      K_contrast_unnorm = K_diff$unnormalized[n_last],
      K_contrast_norm = K_diff$normalized[n_last],
      stringsAsFactors = FALSE
    )
  }
}
summary_df <- do.call(rbind, rows)

num_cols <- c("rho_true", "rho_unnorm", "rho_norm",
              "rho_relerr_unnorm_pct", "rho_relerr_norm_pct",
              "K_contrast_unnorm", "K_contrast_norm")

datatable(summary_df, rownames = FALSE, filter = "top",
          options = list(pageLength = 15, order = list(list(0, "asc"), list(1, "asc")))) |>
  formatRound(columns = num_cols, digits = 3) |>
  formatStyle("rho_relerr_norm_pct",
              backgroundColor = styleInterval(c(-10, 10), c("#f8d7da", "white", "#f8d7da")))
```

## Convergence plots {.tabset}

Each plot shows how the importance-sampling estimate moves as `nSamples` grows, for both the unnormalized and self-normalized weighting schemes. Use the dropdown (top right of each plot) to switch between reweighting scenarios within that group.

```{r convergence-functions}
plot_rho_convergence <- function(group, labels) {
  n_cases <- length(group)
  p <- plot_ly()
  trace_of <- list()
  idx <- 0

  for (i in seq_len(n_cases)) {
    case <- group[[i]]
    p <- p |> add_lines(x = case$nSamples, y = case$res_is$rho_is,
                         name = "unnormalized", legendgroup = "unnorm",
                         line = list(color = "firebrick", dash = "dot"),
                         visible = (i == 1))
    idx <- idx + 1; trace_of[[paste0(i, "_u")]] <- idx

    p <- p |> add_lines(x = case$nSamples, y = case$res_is$rho_is_norm,
                         name = "normalized", legendgroup = "norm",
                         line = list(color = "steelblue"),
                         visible = (i == 1))
    idx <- idx + 1; trace_of[[paste0(i, "_n")]] <- idx
  }

  rho_true <- group[[1]]$res_true$rho_true
  x_range <- range(unlist(lapply(group, function(g) g$nSamples)))
  p <- p |> add_lines(x = x_range, y = c(rho_true, rho_true), name = "rho_true",
                       line = list(color = "black", dash = "dash"), visible = TRUE)
  idx <- idx + 1
  n_total <- idx

  buttons <- lapply(seq_len(n_cases), function(i) {
    vis <- rep(FALSE, n_total)
    vis[trace_of[[paste0(i, "_u")]]] <- TRUE
    vis[trace_of[[paste0(i, "_n")]]] <- TRUE
    vis[n_total] <- TRUE
    list(method = "restyle", args = list("visible", vis), label = labels[i])
  })

  p |> layout(
    xaxis = list(title = "nSamples", type = "log"),
    yaxis = list(title = "intensity estimate (rho)"),
    updatemenus = list(list(active = 0, x = 1, y = 1.12, xanchor = "right", buttons = buttons))
  )
}

plot_K_convergence <- function(group, labels) {
  n_cases <- length(group)
  p <- plot_ly()
  trace_of <- list()
  idx <- 0

  for (i in seq_len(n_cases)) {
    case <- group[[i]]
    K_diff <- calc_K_diff(K_true = case$res_true$K_true,
                           K_norm = case$res_is$K_is_norm,
                           K_unnorm = case$res_is$K_is)

    p <- p |> add_lines(x = case$nSamples, y = K_diff$unnormalized,
                         name = "unnormalized", legendgroup = "unnorm",
                         line = list(color = "firebrick", dash = "dot"),
                         visible = (i == 1))
    idx <- idx + 1; trace_of[[paste0(i, "_u")]] <- idx

    p <- p |> add_lines(x = case$nSamples, y = K_diff$normalized,
                         name = "normalized", legendgroup = "norm",
                         line = list(color = "steelblue"),
                         visible = (i == 1))
    idx <- idx + 1; trace_of[[paste0(i, "_n")]] <- idx
  }
  n_total <- idx

  buttons <- lapply(seq_len(n_cases), function(i) {
    vis <- rep(FALSE, n_total)
    vis[trace_of[[paste0(i, "_u")]]] <- TRUE
    vis[trace_of[[paste0(i, "_n")]]] <- TRUE
    list(method = "restyle", args = list("visible", vis), label = labels[i])
  })

  p |> layout(
    xaxis = list(title = "nSamples", type = "log"),
    yaxis = list(title = "K-function contrast (lower = closer to truth)"),
    updatemenus = list(list(active = 0, x = 1, y = 1.12, xanchor = "right", buttons = buttons))
  )
}
```

```{r convergence-plots, results = "asis"}
# print()ing a htmlwidget manually inside a results='asis' loop doesn't reliably
# register its JS/CSS dependencies with knitr on its own (tested: the dependency
# is present on the widget object itself, but gets dropped between print() and
# the final document) - knit_meta_add() explicitly registers it so the plotly
# library actually ends up in the rendered HTML.
for (g in seq_along(simStudyResults)) {
  group <- simStudyResults[[g]]
  labels <- vapply(group, describe_case, character(1))

  cat(sprintf("\n### Group %d: %s {.tabset}\n\n", g, format_params(group[[1]]$par_goal)))

  cat("\n#### Intensity (rho)\n\n")
  rho_plot <- plot_rho_convergence(group, labels)
  knitr::knit_meta_add(htmltools::findDependencies(rho_plot))
  print(htmltools::tagList(rho_plot))

  cat("\n#### K-function contrast\n\n")
  K_plot <- plot_K_convergence(group, labels)
  knitr::knit_meta_add(htmltools::findDependencies(K_plot))
  print(htmltools::tagList(K_plot))
}
```
